Online Supplement: \\ How important are functional and developmental constraints on phenotypic evolution? An empirical test with the stomatal anatomy of flowering plants, \\ \textit{The~American~Naturalist}

### 2023-09-06 ms/online-supplement.Rmd
final tweaks after publication

\begin{table}
\caption{\label{tab:pair_div}Final data set of `r n_pair` taxon pairs for analysis. $\tt{tree\_node}$ is the node of the common ancestor of the taxon pair $\tt{sp1}$ and $\tt{sp2}$ in the phylogeny (Notes S3). $\tt{pair\_age}$ is the time in millions of years since taxa split. The remaining columns are the trait divergence (log-scale) between taxa ($\Delta \text{log(trait)}$).}
\end{table}

@@

## Notes S1: Theory connecting developmental function, constraint, and integration

@@

## Notes S2: Phylogeny

@@

We used the R packages **taxonlookup** version `r packageVersion("taxonlookup")` [@pennell_simple_2016] and **V.phylomaker** version `r packageVersion("V.phylomaker")` [@jin_vphylomaker_2019] to maximize overlap between our data set and the GBOTB.extended mega-tree of seed plants [@smith_constructing_2018; @zanne_three_2014]. We further resolved large ($\ge4$ taxa) polytomies in `r nrow(polytomies) - 1` clades with sufficient sequence data using **PyPHLAWD** version 1.0 [@smith_pyphlawd_2019] in `r system("python3 --version", intern = TRUE)` (Python Software Foundation, https://www.python.org/). We used sequence data from the most recent GenBank Plant and Fungal sequences database division [@ouellette_database_1997]. We inferred subtree phylogenies using `r str_extract(str_c(system("raxmlHPC-PTHREADS-SSE3 -version", intern = TRUE), collapse = ""), "RAxML version [0-9]*.[0-9]*.[0-9]*")` [@stamatakis_raxml_2014] and conducted molecular dating using the `chronos()` function in the R package **ape** version `r packageVersion("ape")` [@paradis_ape_2019] to obtain ultrametric trees. We grafted resolved, ultrametric subtrees onto the mega-tree at the polytomy nodes and rescaled to keep the mega-tree ultrametric. In some cases, resolving polytomies was not possible because there was little or no overlap between taxa in the data set and taxa with sequence data available for **PyPHLAWD**. In these cases, we randomly selected two taxa as a phylogenetially independent pair and dropped the rest. Remaining polytomies of three taxa were resolved randomly using the `multi2di()` function in **ape**. The final data set for which we had both trait and phylogenetic information contained `r n_tax3` taxa (Notes S3).

@@

# References Cited Only in the Online Enhancements

>>>>

\begin{table}
\caption{\label{tab:pair_div}Final data set of `r n_pair` taxon pairs for analysis. $\tt{tree\_node}$ is the node of the common ancestor of the taxon pair $\tt{sp1}$ and $\tt{sp2}$ in the phylogeny. $\tt{pair\_age}$ is the time in millions of years since taxa split. The remaining columns are the trait divergence (log-scale) between taxa ($\Delta \text{log(trait)}$).}
\end{table}

@@

## Theory connecting developmental function, constraint, and integration

@@

## Phylogeny

@@

We used the R packages **taxonlookup** version `r packageVersion("taxonlookup")` [@pennell_simple_2016] and **V.phylomaker** version `r packageVersion("V.phylomaker")` [@jin_vphylomaker_2019] to maximize overlap between our data set and the GBOTB.extended mega-tree of seed plants [@smith_constructing_2018; @zanne_three_2014]. We further resolved large ($\ge4$ taxa) polytomies in `r nrow(polytomies) - 1` clades with sufficient sequence data using **PyPHLAWD** version 1.0 [@smith_pyphlawd_2019] in `r system("python3 --version", intern = TRUE)` (Python Software Foundation, https://www.python.org/). We used sequence data from the most recent GenBank Plant and Fungal sequences database division [@ouellette_database_1997]. We inferred subtree phylogenies using `r str_extract(str_c(system("raxmlHPC-PTHREADS-SSE3 -version", intern = TRUE), collapse = ""), "RAxML version [0-9]*.[0-9]*.[0-9]*")` [@stamatakis_raxml_2014] and conducted molecular dating using the `chronos()` function in the R package **ape** version `r packageVersion("ape")` [@paradis_ape_2019] to obtain ultrametric trees. We grafted resolved, ultrametric subtrees onto the mega-tree at the polytomy nodes and rescaled to keep the mega-tree ultrametric. In some cases, resolving polytomies was not possible because there was little or no overlap between taxa in the data set and taxa with sequence data available for **PyPHLAWD**. In these cases, we randomly selected two taxa as a phylogenetially independent pair and dropped the rest. Remaining polytomies of three taxa were resolved randomly using the `multi2di()` function in **ape**. The final data set for which we had both trait and phylogenetic information contained `r n_tax3` taxa. This phylogeny is deposited on Dryad [@muir_data_2022].

@@

# Literature Cited

### 2023-09-06 ms/online-supplement.tex
final tweaks after publication

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
% \documentclass[11pt]{article}
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{titlesec}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{multirow} % for merging rows in tables
\usepackage{caption} % for \captionsetup{labelformat=empty}
\usepackage{nopageno}

@@

@@

\begin{table}
\caption{\label{tab:pair_div}Final data set of 236 taxon pairs for analysis. $\tt{tree\_node}$ is the node of the common ancestor of the taxon pair $\tt{sp1}$ and $\tt{sp2}$ in the phylogeny (Notes S3). $\tt{pair\_age}$ is the time in millions of years since taxa split. The remaining columns are the trait divergence (log-scale) between taxa ($\Delta \text{log(trait)}$).}
\end{table}

@@

\hypertarget{notes-s1-theory-connecting-developmental-function-constraint-and-integration}{%
\subsection{Notes S1: Theory connecting developmental function, constraint, and integration}\label{notes-s1-theory-connecting-developmental-function-constraint-and-integration}}

@@

\hypertarget{notes-s2-phylogeny}{%
\subsection{Notes S2: Phylogeny}\label{notes-s2-phylogeny}}

We resolved taxonomic names using the R package \textbf{taxize} version 0.9.100 (\protect\hyperlink{ref-chamberlain_taxize_2013}{Chamberlain and Szöcs 2013}). We queried taxonomic names supplied by the original study authors on 2022-10-05 from the following sources: GRIN Taxonomy for Plants (\protect\hyperlink{ref-united_states_department_of_agriculture_agricultural_research_service_germplasm_2020}{United States Department of Agriculture, Agricultural Research Service 2020}), Open Tree of Life Reference Taxonomy (\protect\hyperlink{ref-rees_automated_2017}{Rees and Cranston 2017}), The International Plant Names Index (\protect\hyperlink{ref-the_royal_botanic_gardens_international_2020}{The Royal Botanic Gardens et al. 2020}), Tropicos - Missouri Botanical Garden (\protect\hyperlink{ref-missouri_botanical_garden_tropicos_2020}{Missouri Botanical Garden 2020}). We retained the maximum scoring matched name with taxize score \(\ge 0.75\) (a score of 1 is a perfect match). In 5 ambiguous cases we manually curated names. Taxonomic name resolution reduced the data set from 1120 to 1080 taxa. Most taxa are different species, but some recognized subspecies and varieties are also included. All algorithms and choices are documented in the associated source code.

We used the R packages \textbf{taxonlookup} version 1.1.5 (\protect\hyperlink{ref-pennell_simple_2016}{Pennell, FitzJohn, and Cornwell 2016}) and \textbf{V.phylomaker} version 0.1.0 (\protect\hyperlink{ref-jin_vphylomaker_2019}{Jin and Qian 2019}) to maximize overlap between our data set and the GBOTB.extended mega-tree of seed plants (\protect\hyperlink{ref-smith_constructing_2018}{Smith and Brown 2018}; \protect\hyperlink{ref-zanne_three_2014}{Zanne et al. 2014}). We further resolved large (\(\ge4\) taxa) polytomies in 29 clades with sufficient sequence data using \textbf{PyPHLAWD} version 1.0 (\protect\hyperlink{ref-smith_pyphlawd_2019}{Smith and Walker 2019}) in Python 3.9.2 (Python Software Foundation, \url{https://www.python.org/}). We used sequence data from the most recent GenBank Plant and Fungal sequences database division (\protect\hyperlink{ref-ouellette_database_1997}{Ouellette and Boguski 1997}). We inferred subtree phylogenies using RAxML version 8.2.12 (\protect\hyperlink{ref-stamatakis_raxml_2014}{Stamatakis 2014}) and conducted molecular dating using the \texttt{chronos()} function in the R package \textbf{ape} version 5.6.2 (\protect\hyperlink{ref-paradis_ape_2019}{Paradis and Schliep 2019}) to obtain ultrametric trees. We grafted resolved, ultrametric subtrees onto the mega-tree at the polytomy nodes and rescaled to keep the mega-tree ultrametric. In some cases, resolving polytomies was not possible because there was little or no overlap between taxa in the data set and taxa with sequence data available for \textbf{PyPHLAWD}. In these cases, we randomly selected two taxa as a phylogenetially independent pair and dropped the rest. Remaining polytomies of three taxa were resolved randomly using the \texttt{multi2di()} function in \textbf{ape}. The final data set for which we had both trait and phylogenetic information contained 638 taxa (Notes S3).

@@

\hypertarget{references-cited-only-in-the-online-enhancements}{%
\section*{References Cited Only in the Online Enhancements}\label{references-cited-only-in-the-online-enhancements}}
\addcontentsline{toc}{section}{References Cited Only in the Online Enhancements}

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
% \documentclass[11pt]{article}
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{titlesec}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{multirow} % for merging rows in tables
\usepackage{caption} % for \captionsetup{labelformat=empty}
\usepackage{nopageno} % suppress page numbers

@@

\author{}
\date{}

@@

\begin{table}
\caption{\label{tab:pair_div}Final data set of 236 taxon pairs for analysis. $\tt{tree\_node}$ is the node of the common ancestor of the taxon pair $\tt{sp1}$ and $\tt{sp2}$ in the phylogeny. $\tt{pair\_age}$ is the time in millions of years since taxa split. The remaining columns are the trait divergence (log-scale) between taxa ($\Delta \text{log(trait)}$).}
\end{table}

@@

\hypertarget{theory-connecting-developmental-function-constraint-and-integration}{%
\subsection{Theory connecting developmental function, constraint, and integration}\label{theory-connecting-developmental-function-constraint-and-integration}}

@@

\hypertarget{phylogeny}{%
\subsection{Phylogeny}\label{phylogeny}}

We resolved taxonomic names using the R package \textbf{taxize} version 0.9.100 (\protect\hyperlink{ref-chamberlain_taxize_2013}{Chamberlain and Szöcs 2013}). We queried taxonomic names supplied by the original study authors on 2022-12-01 from the following sources: GRIN Taxonomy for Plants (\protect\hyperlink{ref-united_states_department_of_agriculture_agricultural_research_service_germplasm_2020}{United States Department of Agriculture, Agricultural Research Service 2020}), Open Tree of Life Reference Taxonomy (\protect\hyperlink{ref-rees_automated_2017}{Rees and Cranston 2017}), The International Plant Names Index (\protect\hyperlink{ref-the_royal_botanic_gardens_international_2020}{The Royal Botanic Gardens et al. 2020}), Tropicos - Missouri Botanical Garden (\protect\hyperlink{ref-missouri_botanical_garden_tropicos_2020}{Missouri Botanical Garden 2020}). We retained the maximum scoring matched name with taxize score \(\ge 0.75\) (a score of 1 is a perfect match). In 5 ambiguous cases we manually curated names. Taxonomic name resolution reduced the data set from 1120 to 1080 taxa. Most taxa are different species, but some recognized subspecies and varieties are also included. All algorithms and choices are documented in the associated source code.

We used the R packages \textbf{taxonlookup} version 1.1.5 (\protect\hyperlink{ref-pennell_simple_2016}{Pennell, FitzJohn, and Cornwell 2016}) and \textbf{V.phylomaker} version 0.1.0 (\protect\hyperlink{ref-jin_vphylomaker_2019}{Jin and Qian 2019}) to maximize overlap between our data set and the GBOTB.extended mega-tree of seed plants (\protect\hyperlink{ref-smith_constructing_2018}{Smith and Brown 2018}; \protect\hyperlink{ref-zanne_three_2014}{Zanne et al. 2014}). We further resolved large (\(\ge4\) taxa) polytomies in 29 clades with sufficient sequence data using \textbf{PyPHLAWD} version 1.0 (\protect\hyperlink{ref-smith_pyphlawd_2019}{Smith and Walker 2019}) in Python 3.9.2 (Python Software Foundation, \url{https://www.python.org/}). We used sequence data from the most recent GenBank Plant and Fungal sequences database division (\protect\hyperlink{ref-ouellette_database_1997}{Ouellette and Boguski 1997}). We inferred subtree phylogenies using RAxML version 8.2.12 (\protect\hyperlink{ref-stamatakis_raxml_2014}{Stamatakis 2014}) and conducted molecular dating using the \texttt{chronos()} function in the R package \textbf{ape} version 5.6.2 (\protect\hyperlink{ref-paradis_ape_2019}{Paradis and Schliep 2019}) to obtain ultrametric trees. We grafted resolved, ultrametric subtrees onto the mega-tree at the polytomy nodes and rescaled to keep the mega-tree ultrametric. In some cases, resolving polytomies was not possible because there was little or no overlap between taxa in the data set and taxa with sequence data available for \textbf{PyPHLAWD}. In these cases, we randomly selected two taxa as a phylogenetially independent pair and dropped the rest. Remaining polytomies of three taxa were resolved randomly using the \texttt{multi2di()} function in \textbf{ape}. The final data set for which we had both trait and phylogenetic information contained 638 taxa. This phylogeny is deposited on Dryad (\protect\hyperlink{ref-muir_data_2022}{Muir et al. 2022}).

@@

\hypertarget{literature-cited}{%
\section*{Literature Cited}\label{literature-cited}}
\addcontentsline{toc}{section}{Literature Cited}

@@

\leavevmode\vadjust pre{\hypertarget{ref-muir_data_2022}{}}%
Muir, Christopher D, Miquel Àngel Conesa, Jeroni Galmés, Varsha S Pathare, Patricia Rivera, Rosana López Rodríguez, Teresa Terrazas, and Dongliang Xiong. 2022. {``Data from: {How} Important Are Functional and Developmental Constraints on Phenotypic Evolution? {An} Empirical Test with the Stomatal Anatomy of Flowering Plants.''} \url{https://doi.org/10.5061/dryad.0rxwdbs42}.
